---
title: 'CMTH 642 Data Analytics: Advanced Methods Assignment 1'
output:
  html_document: default
  word_document: default
  pdf_document: default
---

Assignment 1 is worth 10% of the final grade. Submit the .rmd file and the generated output file (either in PDF, Word or HTML format). Failing to submit both files will be subject to a mark deduction.

Preparation:

The dataset is related to Forest Fires. For more info:
https://archive.ics.uci.edu/ml/datasets/forest+fires


### 1. Read the csv file from the this URL https://archive.ics.uci.edu/ml/machine-learning-databases/forest-fires/forestfires.csv.   Save it to a data frame called Forest. Check the first six records.  (10 points)
```{r}
# Load required package and read the CSV file
library(readr)
Forest <- read_csv("https://archive.ics.uci.edu/ml/machine-learning-databases/forest-fires/forestfires.csv")

# First six records of the data frame
head(Forest)
```

### 2. How many observations are there in the dataset? How many observations are there with both a fire (burned area > 0) and rain? (10 points)
```{r}
# Number of observations in the data frame
nrow(Forest)

# Create a subset of the data frame that only contains rows where the burned area and where the rain is greater than 0
Forest_burned <- Forest[Forest$area > 0, ]

Forest_rain <- Forest[Forest$rain > 0, ]

# Number of observations where both conditions are true
nrow(intersect(Forest_burned, Forest_rain))
```

### 3. Show the columns month, day, area of all the observations with a fire (burned area > 0). (10 points)
```{r}
# Create subset
Forest_burned <- Forest[Forest$area > 0, ]

# Select the columns of interest
Forest_burned[, c("month", "day", "area")]

```

### 4. How large are the five largest fires (i.e. having largest area)? (10 points)
```{r}
library(dplyr)

# Sort the data frame by area in descending order and select the top five rows
largest_fires <- head(arrange(Forest, desc(area)), 5)

largest_fires$area

```

### 5. For the records you obtained from the previous question, what are the corresponding month, temp, RH, wind, rain, area? (10 points)
```{r}
largest_fires[, c("month", "temp", "RH", "wind", "rain", "area")]
```

### 6. For the whole dataset, reorder factor levels of month to be from Jan to Dec. (10 points)
```{r}
# Reorder the factor levels of the month column
Forest$month <- factor(Forest$month, levels = c("jan", "feb", "mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec"))
levels(Forest$month)

```
### 7. Add one column to the dataframe indicating whether a fire occurred for each observation ('TRUE' for area>0 and 'FALSE' for area==0). (10 points)
```{r}
# Create a new column indicating whether a fire occurred
Forest$fire_occurred <- ifelse(Forest$area > 0, TRUE, FALSE)
head(Forest)

```
### 8. What is the mean area/wind/temp/RH per month? (10 points)
```{r}
# Group the data by month and calculate the mean area, wind, temp, and RH
Forest %>%
  group_by(month) %>%
  summarize(mean_area = mean(area),
            mean_wind = mean(wind),
            mean_temp = mean(temp),
            mean_RH = mean(RH))

```

### 9. How many observations are there in each month? Visualize by using a bar plot in descending order. (10 points)
```{r}
library(dplyr)
library(ggplot2)

# Define colors of the bars
colors <- c("#619CFF", "#00D166", "#FF7F00", "#FF006E", "#B15AED", "#6A4C93", 
               "#FFC100", "#6DD9EC", "#FF006E", "#00D166", "#619CFF", "#B15AED")

# Count observations by month
count_df <- Forest %>% 
  group_by(month) %>% 
  summarize(n = n())

# Create the plot
ggplot(count_df, aes(x = reorder(month, -n), y = n, fill = month)) +
  geom_bar(stat = "identity") +
  labs(x = "Month", y = "Number of observations") +
  theme_bw() +
  scale_fill_manual(values = colors)



```

### 10. How many observations are there with burned area greater than zero in each month? Visualize by using a bar plot in descending order. (10 points)
```{r}

# Define color palette
colors <- c("#619CFF", "#00D166", "#FF7F00", "#FF006E", "#B15AED", "#6A4C93", 
               "#FFC100", "#6DD9EC", "#FF006E", "#00D166", "#619CFF", "#B15AED")

# Count observations with area > 0 by month
count_df <- Forest %>% 
  filter(area > 0) %>% 
  group_by(month) %>% 
  summarize(n = n())

# Create the plot with custom colors
ggplot(count_df, aes(x = reorder(month, -n), y = n, fill = month)) +
  geom_bar(stat = "identity") +
  labs(x = "Month", y = "Observations with area > 0") +
  theme_bw() +
  scale_fill_manual(values = colors)

```

This is the end of Assignment 1 

Ceni Babaoglu, PhD
